---
title: "Introduction to R"
subtitle: "IQA Lecture 0"
author: "Charles Lanfear"
date: "6 Oct 2022<br>Updated: `r gsub(' 0', ' ', format(Sys.Date(), format='%d %b %Y'))`"
output:
  xaringan::moon_reader:
    css: [xaringan-themer.css, "../assets/cam_style.css"]
    lib_dir: libs
    nature:
      highlightStyle: tomorrow-night-bright
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "../assets/cam_macros.js"
      titleSlideClass: ["center","top"]
---

```{r setup}
#| include: false
knitr::opts_chunk$set(eval=TRUE, echo=TRUE, message=TRUE, warning=TRUE)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  white_color = "#f0f1eb",
  black_color = "#28282B",
  primary_color = "#003C71",
  secondary_color = "#8A1538",
  header_font_google = google_font("Montserrat"),
  text_font_google   = google_font("Open Sans"),
  code_font_google   = google_font("Fira Mono"),
  padding = "16px 32px 16px 32px",
  text_font_size = "1.3rem",
  code_font_size = "1.1rem"
)
```


# Course Goals


* Cultivate basic competence with applied statistics

  * *We'll start this next week*

--

* Develop intermediate R data management and visualization skills

  * *We'll start this today*

--

* Prepare you for advanced topics in quantitative analysis

--

* Introduce reproducible research practices

---
class: inverse

.text-align-center[
# Why are you doing this to me?
]

---

# Reasons

--

## You *will* read quantitative research

* Interpretation
* Evaluation

--

## You *may* do quantitative research

* Analysis
* Reporting

---
class: inverse

.text-align-center[
# Okay, but programming?
]

---
# Data *analysis* is mostly data *wrangling*<sup>1</sup>

.footnote[Credit: Kieran Healy]

--

&nbsp;

## Data wrangling is **frustrating**

---
class: inverse

.text-align-center[
# Well, can we make it fun?
]

---

&nbsp;


.text-big-center[
**NO**
]

---
class: inverse

.text-align-center[
# Can we at least make it not frustrating?
]

---

&nbsp;

.text-big-center[
**ALSO<br>NO**
]

---
# Programming is *important*

* Cleaning data by hand is:

  * Error prone
  * Time consuming
  * Not **reproducible**

--

* It opens *many* doors

--

* Frustration is *always* present

--

* But what frustrates you *changes*

---

# Logistics

Location: Computer Lab, Wed, 14:00-16:00

My email: [cl948@cam.ac.uk]()

Drop-in times:

* 3.9, Tue, 14:00-15:00
* 3.9, Wed, 11:00-12:00

Materials: [http://clanfear.github.io/ioc_iqa]()

---

# Materials

All course materials are on the course website. This includes:

* These slides and the code used to generate them.
* An R script for the slides to follow along in class.
* Useful links to other resources.
* **Formative assignments**

If something doesn't work, email me.

---

# A Note on Slide Formatting

**Bold** is an important term.

--

*Italics* are emphasis or mouse-click instructions

* "Click *File \> Print*"

--

`Code` is R code you type in or keystrokes for actions

* "Press `Ctrl-P` to open the print dialogue."

--

Code chunks are *actual embedded R code*.

```{r}
#| code-line-numbers: "1|2"
# Sometimes important stuff is highlighted! #<<
7 * 49
```

--

`##` indicates the output from running the code chunk.

---
class: inverse

# R and RStudio

---

# Why R?

R is a programming language built for statistical computing.

If one already knows Excel, SPSS, or Stata, why use R?

--

* R is *free*, so you don't need a terminal server or license.

--

* R has a *very* large community for support and packages.

--

* R can handle virtually any data format.

--

* R makes replication *easy*.

--

* R is a *language* so it can do *everything*.<sup>1</sup>

--

* R is similar to other programming languages.


---

# R Studio

R Studio is a "front-end" or integrated development environment (IDE) for R that can make your life *easier*.

--

We'll show RStudio can...

* Organize your code, output, and plots

--

* Auto-complete code and highlight syntax

--

* Help view data and objects

--

* Enable easy integration of R code into documents with **Quarto**

--

It can also...

* Manage `git` repositories

* Run interactive tutorials

* Handle other languages like Python, SQL, and HTML

---

# Selling you on Quarto

The ability to create Quarto files is a powerful advantage of R:

* Document analyses by combining text, code, and output

--

* No copying and pasting into Word

--

* Easy for collaborators to understand

--

* Show as little or as much code as you want

--

* Produce many different document types as output

  * PDF documents
  * HTML webpages and reports
  * Word and PowerPoint documents
  * Presentations

--

* Works with LaTeX and HTML for math and formatting

--

We'll get back to this shortly!

---

# Getting Started

Open up RStudio now and choose *File \> New File \> R Script*.

Then, let's get oriented with the interface:

--

* *Top Left*: Code **editor** pane, data viewer (browse with tabs)

--

* *Bottom Left*: **Console** for running code (`>` prompt)

--

* *Top Right*: List of objects in **environment**, code **history** tab.

--

* *Bottom Right*: Tabs for browsing files, viewing plots, managing packages, and viewing help files.

--

You can change the layout in *Preferences \> Pane Layout*

---

# Editing and Running Code

There are several ways to run R code in RStudio:

--

* Highlight lines in the **editor** window and click *Run* at the top or hit `Ctrl+Enter` or `⌘+Enter` to run them all.

--

* With your **caret** (.blink[`|`]) on a line you want to run, hit `Ctrl+Enter` or `⌘+Enter`. Note your caret moves to the next line, so you can run code sequentially with repeated presses.

--

* Type individual lines in the **console** and press `Enter`.

The console will show the lines you ran followed by any printed output.

---

# Incomplete Code

If you mess up (e.g. leave off a parenthesis), R might show a `+` sign prompting you to finish the command:

```{r Coding 1, eval=FALSE}
> (11-2
+
```

Finish the command or hit `Esc` to get out of this.

---

# R as a Calculator

In the **console**, type `123 + 456 + 789` and hit `Enter`.

--

```{r Calc 1}
123 + 456 + 789
```

--

The `[1]` in the output indicates the numeric **index** of the first element on that line.

--

Now in your blank R document in the **editor**, try typing the line `sqrt(400)` and either clicking *Run* or hitting `Ctrl+Enter` or `⌘+Enter`.

--

```{r Calc 2}
sqrt(400)
```

---
# Functions and Help

`sqrt()` is an example of a **function** in R.

--

If we didn't have a good guess as to what `sqrt()` will do, we can type `?sqrt` in the console and look at the **Help** panel on the right.

```{r Help, eval=FALSE}
?sqrt
```

--

**Arguments** are the *inputs* to a function. In this case, the only argument to `sqrt()` is `x` which can be a number or a vector of numbers.

Help files provide documentation on how to use functions and what functions produce.

---

# Creating Objects

R stores *everything* as an **object**, including data, functions, models, and output.

--

Creating an object can be done using the **assignment operator**: `<-`

--

```{r Objects 1}
new.object <- 144
```

--

**Operators** like `<-` are functions that look like symbols but typically sit between their arguments (e.g. numbers or objects) instead of having them inside `()` like in `sqrt(x)`.<sup>1</sup>

.footnote[[1] We can call operators like other functions using backticks: <code>\`+\`(x,y)</code>]

--

We do math with operators, e.g., `x + y`. `+` is the addition operator!

---

# Calling Objects

You can display or "call" an object simply by using its name.

```{r Objects 2}
new.object
```

--

Object names can contain `_` and `.` in them, but cannot *begin* with numbers. Try to be consistent in naming objects.

--

Auto-complete means *long names are better than vague ones*!

*Good names*<sup>1</sup> save confusion later.

.footnote[[1] *"There are only two hard things in Computer Science: cache invalidation and naming things." - Phil Karlton*]

---

# Using Objects

You can treat the object's **name** as if it were the values stored inside.

--

```{r Objects 3}
new.object + 10
new.object + new.object
sqrt(new.object)
```

---

# Creating Vectors

A **vector** is a series of **elements**, such as numbers.

--

You can create a vector using the function `c()` which stands for "combine".

--

```{r Vectors 1}
new.object <- c(4, 9, 16, 25, 36)
new.object
```

--

Assigning to an existing name *overwrites that object*.

--

You can provide a vector as an argument for many functions.

--

```{r Vectors 2}
sqrt(new.object)
```

---

# More Complex Objects

The same principles can be used to create more complex objects:

* **matrices**
* **arrays**
* **lists**
* **dataframes**

--

Most data sets you will work with will be read into R and stored as a **dataframe**. 

Consequently, we will focus on manipulating and analyzing these dataframes.

--

Before we get into these, let's revisit **Quarto**.

---
class: inverse
# Quarto

![](img/quarto.svg)

---

## Quarto Documents

Let's try making a Quarto file:

1.  Choose *File \> New File \> Quarto Document...*
2.  Make sure *HTML* is selected and click OK
3.  Save the file somewhere, call it `my_first_qmd.qmd`
4.  Click the *Render* button
5.  Watch the progress in the jobs pane, then gaze upon your result!

You may also open up the file in your computer's browser if you so desire, using the *Open in Browser* button at the top of the preview window.

---

# YAML Frontmatter

The frontmatter (or header) of a `.qmd` file is a [YAML](http://yaml.org/) code block, and everything else is part of the main document.


```
---
title: "Untitled"
author: "Your name"
format: html
editor: visual
---
```

--

To change how the document looks, [you can modify the header with themes](https://quarto.org/docs/output-formats/html-themes.html).<sup>1</sup>


```
format: 
  html: 
    theme: darkly
```

---

# Markdown Syntax

.pull-left[
## Output

**bold/strong emphasis**

*italic/normal emphasis*

# Header

## Subheader

### Subsubheader

> Block quote from famous person

]

.pull-right[
## Syntax

<pre>
**bold/strong emphasis**
*italic/normal emphasis*

# Header

## Subheader

### Subsubheader

> Block quote from
> famous person
</pre>
]

---

# More Markdown Syntax

.pull-left[
## Output

1.  Ordered lists
2.  Are real easy
  3.  Even with sublists
  4.  Or when lazy with numbering

* Unordered lists
* Are also real easy
    * Also even with sublists

[URLs are trivial](http://www.uw.edu)

![:width 200px](https://www.cam.ac.uk/sites/www.cam.ac.uk/themes/fresh/images/interface/cambridge_university2.svg)
]

.pull-right[
## Syntax

<pre>
1. Ordered lists
1. Are real easy
  1. Even with sublists
  1. Or when lazy with numbering


* Unordered lists
* Are also real easy
  + Also even with sublists

[URLs are trivial](http://www.uw.edu)
![Images too](https://www.cam.ac.uk/sites/www.cam.ac.uk/themes/fresh/images/interface/cambridge_university2.svg)
</pre>
]

---

# Formulae and Syntax

.pull-left[
### Output

You can put some math $y= \left( \frac{2}{3} \right)^2$ right up in there.

$$\frac{1}{n} \sum_{i=1}^{n} x_i = \bar{x}_n$$

Or a sentence with `code-looking font`.

Or a block of code:

```
y <- 1:5
z <- y^2

```
]

.pull-right[
### Syntax

```
You can put some math $y= 
\left(\frac{2}{3} \right)^2$ 
right up in there

$$\frac{1}{n} \sum_{i=1}^{n}
x_i = \bar{x}_n$$

Or a sentence with `code-
looking font`.
```

<pre>
Or a block of code:

```
y <- 1:5
z <- y^2
```
</pre>
]

---

# Quarto Tinkering

Quarto docs can be modified in many ways. Visit these links for more information.

* [The Quarto Guide](https://quarto.org/docs/guide/)
* [Ways to modify the overall document appearance](https://quarto.org/docs/output-formats/html-themes.html)

---

# Formatting Caveats

To keep Quarto simple, it lacks some features you might want to use. Your options for fancier documents are:

* Templates
* Use HTML with CSS for custom formatting<sup>5</sup>
* Use custom LaTeX templates<sup>6</sup>

.footnote[
[5] These slides were created using using custom CSS<br>
[6] Here be dragons! LaTeX is powerful but exacts a terrible price
]

For day-to-day use, plain vanilla R Markdown does the job.

For handouts, memos, and assignments, default R Markdown PDFs look surprisingly good!

---

# R Code in R Markdown

Inside quarto,yu can place lines of code in **chunks**. This chunk...

````{verbatim, lang = "markdown"}
```{r}
summary(cars)
```
````

Produces this output in your document:

```{r}
summary(cars)
```

---


# Chunk Options

Chunks have options that control what happens with their code, such as:

* `echo: false`: Hides R code in the document

* `eval: false`: Prevents R code from running

* `include: false`: Hides output but still runs code

* `results: 'hide'`: Hides non-plot output

* `cache: true`: Saves results so you won't have to re-run it each time you render the document

* `fig.height: 5, fig.width: 5`: modify the dimensions of any plots that are generated in the chunk (units are in inches)

[See a lot more about chunk options here](https://yihui.name/knitr/options/).

---

# Playing with Chunk Options

Try adding or changing the chunk options (separated by commas) for the two chunks in `my_first_qmd.qmd` and re-knitting to check what happens.

--

You can name chunks by putting a name on the first `#|` line.

````{verbatim, lang = "markdown"}
```{r}
#| summarize_cars
#| echo: false
summary(cars)
```
````

After you name your chunks, look what happens in the dropdown on the bottom left of your editor pane.

--

You can browse through an RMarkdown document by named chunks! 

You can also browse by sections named using headers (`#`) and subheaders (`##`).

---

# In-Line R code

Sometimes we want to insert a value directly into our text. We do that using code in single backticks starting off with `r`.

--

```{r, echo=FALSE}
library(knitr)
```

```
Four score and seven years ago is the same as `r knitr::inline_expr("4*20 + 7")` years.
```

--

Four score and seven years ago is the same as `r 4*20 + 7` years.

--

Maybe we've saved a variable in a chunk we want to reference in the text:

```{r}
x <- sqrt(77) # <- is how we assign objects
```

--

```
The value of `x` rounded to the nearest two decimals is 
`r knitr::inline_expr("round(x, 2)")`.
``` 

--

The value of `x` rounded to the nearest two decimals is `r round(x, 2)`.

---

# This is Amazing!

Having R dump values directly into your document protects you from silly mistakes:

--

* Never wonder "how did I come up with this quantity?" ever again: Just look at your formula in your .qmd file!

--

* Consistency! Reference a variable in-line throughout your document without manually updating if the calculation changes (e.g. reporting sample sizes).

--

* You are more likely to make a typo in a "hard-coded" number than you are to write R code that somehow runs but gives you the wrong thing.

---
class: inverse

# Data Frames

---

# What's Up with `cars`?

In our Quarto doc, let's load the built-in dataframe `cars`. 

--

`data(cars)` loads this dataframe into the **Global Environment** (as a *promise*<sup>1</sup>).

.pull-right[.footnote[[1] Promises are *unevaluated arguments*. Read more about R's [lazy evaluation here](http://adv-r.had.co.nz/Functions.html).]]

--

`View(cars)` pops up a **Viewer** pane ("interactive" use only, don't put in R Markdown document!) or...

--

```{r}
head(cars, 5) # prints first 5 rows, see tail() too
```



---

# Tell Me More About `cars`

`str()` displays the structure of an object:

.text-85[
```{r}
str(cars) # str[ucture]
```
]

--

`summary()` displays summary information<sup>1</sup>:

.text-85[
```{r}
summary(cars)
```
]

.pull-right-40[.footnote[[1] [Note R is **object-oriented**](https://adv-r.hadley.nz/oo.html): `summary()` provides different information for different types of objects!]]

---

# Ugly Pictures of `cars`

`hist()` generates a histogram of a vector. Note you can access a vector that is a column of a dataframe using `$`, the **extract operator**.

.pull-left[
```{r, out.height="330px"}
hist(cars$speed) # Histogram
```
]

.pull-right[
```{r, out.height="330px"}
hist(cars$dist)
```
]

---

## Drawing Slightly Less Ugly Pictures


```{r, out.height="400px"}
hist(cars$dist,
     xlab = "Distance (ft)", # X axis label
     main = "Observed stopping distances of cars") # Title
```


---

# Math with `cars`

If you put an assignment such as `x <- y` in parentheses `()`, R will print the output of the assignment out for you in your document. Otherwise, it won't show the value.
--

```{r}
( dist_mean  <- mean(cars$dist) )
( speed_mean <- mean(cars$speed) )
```

---

# Drawing Still Ugly Pictures

.text-72[
```{r, out.height="330px"}
plot(dist ~ speed, data = cars, #<<
     xlab = "Speed (mph)",
     ylab = "Stopping distance (ft)",
     main = "Speeds and stopping distances of cars",
     pch = 16) # Point size
abline(h = dist_mean, col = "firebrick")
abline(v = speed_mean, col = "cornflowerblue")
```
]

.pull-right[
.footnote[Note that `dist ~ speed` is a **formula** of the type `y ~ x`. The first element (`dist`) gets plotted on the y-axis and the second (`speed`) goes on the x-axis. Regression formulae follow this convention as well!]
]

---
# Installing Packages

Let's make a table that looks a little less code-y in the output. To do this, we'll 
want to install a **package** called `pander`. Packages contain premade functions 
and/or data we can use. R's strength is its wide variety of packages!

In the console: `install.packages("pander")`.

--

* Note that unlike the `library()` command, *the name of a package to be installed must be in quotes*. This is because the name here is a search term (text, not an object!) while for `library()` it is an actual R object.

--

* Once you install a package, you don't need to re-install it until you update R. *Consequently, you should not include* `install.packages()` *in any markdown document or R script!*

---

# Making Tables

```{r, eval=FALSE}
library(pander) # loads pander, do once in your session
pander(summary(cars), style = "rmarkdown") #<<
```

.text-72[
```{r, echo=FALSE}
library(pander) # loads pander, do once in your session
pander(summary(cars), style = "rmarkdown")
```
]

.footnote[Note that we put the `summary(cars)` function call inside the `pander()` call. This is called *nesting functions*. I'll introduce a method next week to simplify this!]


---
class: inverse

# For Next Time

* Read first two chapters of Kaplan (2022)
   * Try to run the code in Chapter 2 as you read<sup>1</sup>

* Spend 30-60 minutes playing with your `.qmd` file
   * Use the `cars` data to calculate basic numbers (e.g., `mean()`)
   * Try some in-line calculations
   * Make a table and/or a plot
   
.footnote[[1] If you run into problems you can't get past quickly, set it aside until next week!]
