---
title: "Linear Models IV"
subtitle: "IQA Lecture 7"
author: "Charles Lanfear"
date: "14 Nov 2022<br>Updated: `r gsub(' 0', ' ', format(Sys.Date(), format='%d %b %Y'))`"
output:
  xaringan::moon_reader:
    css: "../assets/cam-css.css"
    lib_dir: libs
    nature:
      highlightStyle: tomorrow-night-bright
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "../assets/cam_macros.js"
      titleSlideClass: ["center","top"]
---

```{r setup, purl=FALSE}
#| include: false
options(width = 68)
set.seed(7)
knitr::opts_chunk$set(eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, dev = "svg", fig.height = 4)
```


# Today

### Comparing Models

### Pivoting Data

### Polynomials

### Residual Plots


---

# Setup

Let's pick up where we left off last week with `metro_2021`

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(broom) 
metro_2021 <- read_csv("https://clanfear.github.io/ioc_iqa/_data/metro_2021_full.csv") |>
  rename(violence = violence_and_sexual_offences,
         asb      = antisocial_behaviour)
```

---


```{r}
lm_viol_2 <- lm(violence ~ month + I(month^2), data = metro_long)
```

---
class: inverse

# Wrap-Up

Reading:

* Huntington-Klein, N. (2022) *The Effect: An Introduction to Research Design and Causality*, New York, NY: Chapman and Hall/CRC Press. 
   * Read the rest of Chapter 13: Regression

* Kaplan, J. (2022) *Crime by the Numbers: A Criminologist's Guide to R*
   * Read [Chapter 13: Reshaping data](https://crimebythenumbers.com/reshaping.html)