---
title: "Linear Models I"
subtitle: "IQA Lecture 4"
author: "Charles Lanfear"
date: "2 Nov 2022<br>Updated: `r gsub(' 0', ' ', format(Sys.Date(), format='%d %b %Y'))`"
output:
  xaringan::moon_reader:
    css: "../assets/cam-css.css"
    lib_dir: libs
    nature:
      highlightStyle: tomorrow-night-bright
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "../assets/cam_macros.js"
      titleSlideClass: ["center","top"]
---

```{r setup, purl=FALSE}
#| include: false
options(width = 68)
knitr::opts_chunk$set(eval=TRUE, echo=TRUE, message=TRUE, warning=TRUE, dev = "svg")
```


# Things to include

How do multiple vars work?

* Demeaning with dummies
* Partial correlations (is that pairs plot?)

* Stats
   * TTest is an lm() with 2 category (binary) predictor
   * ANOVA is an lm() with 3+ category predictor

* Code


---
class: inverse

# Controls

---


```{r}
library(tidyverse)
communities <- 
  read_csv("https://clanfear.github.io/ioc_iqa/_data/communities.csv") |>
  mutate(across(c(incarceration, disadvantage), ~ factor(., levels = c("Low", "Medium", "High"))),
         area = factor(area, levels = c("Rural", "Urban")))
```

---

# Dummy Variables

What is a regression model doing when you have a binary predictor?

--

.pull-left[

```{r}
communities |>
  group_by(area) |>
  summarize(crime_rate = mean(crime_rate))
```
We can calculate group-specific average crime rates


]
.pull-right[

```{r}
dummy_model <- lm(crime_rate ~ area, data = communities)
dummy_model
```

The intercept is the average crime rate for rural places

The value for `areaUrban` is the difference between rural and urban places

]

Linear regression is just estimating means by group!

---

# Testing


```{r}
summary(dummy_model)
```

--

This is what a T-test is!

```{r}
t.test(crime_rate ~ area, data = communities, var.equal=TRUE)
```


---

# More Categories

When you have 3+ 

```{r}
cat_lm <- lm(crime_rate ~ incarceration, data = communities)
cat_lm

```

```{r}
communities |>
  group_by(incarceration) |>
  summarize(crime_rate = mean(crime_rate))
```

---

# ANOVA

```{r}
summary(cat_lm)
```



```{r}
summary(aov(crime_rate ~ incarceration, data = communities))
```

```{r}
coef(aov(crime_rate ~ incarceration, data = communities))
```


---

# Dummy Controls

Some of X and Y's relationship is due to groups having different average levels of both X and Y

If we remove group-specific averages of X and Y, we can get rid of that

```{r}
lm(crime_rate ~ pop_density, data = communities)
```


```{r}
communities |>
  group_by(area) |>
  mutate(pop_density_res = pop_density - mean(pop_density),
         crime_rate_res  = crime_rate - mean(crime_rate)) |>
  lm(crime_rate_res ~ pop_density_res, data = _)
```

```{r}
lm(crime_rate ~ pop_density + area, data = communities)
```


---

# Continuous Variables

Some of X and Y's relationship is because Z makes X and Y higher

If we remove variation in X and Y due to Z, we can get rid of that

This is a bit more complicated—to get the residuals, we need to run bivariate models predicting what we want to residualize

```{r}
communities %>%
  mutate(area_num       = as.numeric(area == "Urban"),
         crime_rate_res = residuals(lm(crime_rate ~ pop_density, data = .)),
         area_resid     = residuals(lm(area_num ~ pop_density, data = .))) |>
  lm(crime_rate_res ~ area_resid, data = _)
```



---


# Curves

```{r}
form <- y ~ x + I(x^2)
communities %>%
  ggplot(aes(x = `Pop Density`, y = `Crime Rate`)) + 
  geom_point(alpha = 0.5) +
  xlab("Population Density") +
  stat_poly_line(formula = form, se = FALSE) +
  stat_poly_eq(formula = form, mapping = use_label("eq"), eq.x.rhs = "PopDensity", eq.with.lhs= "CrimeRate~`=`~") +
  theme_minimal(base_size = 16)
```


---

# Collider Bias


```{r, fig.height = 3.5}
ex_data <- tibble(Skill = runif(1000, 0, 10), 
                  Frequency = runif(1000, 0, 10), 
                  Who = c(rep("Arrested", 500), 
                          rep("Everyone", 500))) |>
  filter(Who == "Everyone" | (10 + (-1*Skill) + Frequency > 10))

ex_data |> 
  ggplot(aes(x = Skill, y = Frequency, color = Who)) + 
  geom_point() + 
  labs(x = "Skill at Crime", y = "Frequency of Crime") +
  geom_smooth(method = "lm", formula = "y~x", se = FALSE, color = "black") + 
  facet_wrap(~Who)  + 
  theme_minimal(base_size = 16) + 
  theme(legend.position = "none",
        panel.spacing.x = unit(0.4, "in"))
```

