---
title: "Linear Models III"
subtitle: "IQA Lecture 6"
author: "Charles Lanfear"
date: "14 Nov 2022<br>Updated: `r gsub(' 0', ' ', format(Sys.Date(), format='%d %b %Y'))`"
output:
  xaringan::moon_reader:
    css: "../assets/cam-css.css"
    lib_dir: libs
    nature:
      highlightStyle: tomorrow-night-bright
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "../assets/cam_macros.js"
      titleSlideClass: ["center","top"]
---

```{r setup, purl=FALSE}
#| include: false
options(width = 68)
set.seed(7)
knitr::opts_chunk$set(eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, dev = "svg", fig.height = 4)
```


# Today

Comparing Models

Functional Forms

* Logs
* Polynomials

Residual Plots

* Non-linearities
* QQ Plots

---

# Specification Tests

**Specification tests** are hypothesis tests for evaluating how well a model fits

--

They can be used to test if...

* One model fits better than another

--

* Coefficients are equal to a value (e.g., 0) or to each other

   * The t-test in regression output is doing this!
   * Useful if you're considering *combining* two variables

--

* Many other assumptions are violated

--

&nbsp;

.text-center[
*We'll do all of these things by the end of the term, but for now, let's focus on *comparing models**
]

---

# `anova()` for Comparisons

The `anova()` function can be used to compare linear models

```{r}
anova(lm_2, lm_3)
```


This produces a chi-square test of whether a given specification significantly reduces the sum of squares—that is, whether it fits better

--

Here `area` does *not* statistically significantly improve the model

--

This statistical test requires one model be **nested** in the other

---

# Nested Models

Models are nested if every variable in the *less complex* model is included in the *more complex model*

--

When models are nested, we say...

* The less complex one is the **restricted model**
* The more complex one is the **unrestricted model**

--

*What exactly is **restricted**?*

--

Excluding a variable from a model is equivalent to settings its *coefficient to zero*

* i.e., we *restrict the coefficient to zero* when we omit that variable

--

When we exclude a path from a DAG, we *assume a coefficient of zero for that path*

* We can test this assumption using `anova()` and other specification tests
* e.g., if our DAG indicated no direct effect of `area` on `crime_rate`, the prior test would support that assumption

---

# Another Example

```{r}
anova(lm_1, lm_2, lm_3)
```

We can give `anova()` any number of models

* Best to start with the *least complex* or *most restricted*
* Each model is compared to the one directly above it

---

# Warning

Adding a collider to a model will almost always improve fit

* `anova()` only tests if the model fits better
* `anova()` doesn't know about your DAG

--

.text-85[
```{r}
lm_4 <- lm(crime_rate ~ disadvantage + incarceration, data = communities)
anova(lm_1, lm_4)
```
]

--

**Theory** is your first guide for what to include or not

This test is mainly to see if it is okay to exclude *potential confounder*

---
# class: inverse

# Polynomials


---

Bering back metro datat unpivoted

```{r}
metro_wide <- read_csv("https://clanfear.github.io/ioc_iqa/_data/metro_2021_violence_wide.csv")
```


---

Pivoting data




---

Do a violence model by month

---

Plot residials

Residuals should be flat with respect to predictors!

You're failing to account for a curved relationship; can't remove a curve by subtracting a straight line

---

Show geom_smooth loess, see the curve


---

Introduce a quadratic polynomial


---

Run the model


---

Plot the residuials


---

Interpet the model


---

Bring back the calculus from first week


---

ggplot(metro_2021, aes(x = month, y = violence_and_sexual_offences)) + geom_point() + geom_smooth()

ggplot(metro_2021, aes(x = month, y = violence_and_sexual_offences)) + geom_point() + geom_smooth(method = "lm", formula = y ~ x + I(x^2))

fixest::feols(violence_and_sexual_offences ~ month + I(month^2) | borough, data =  metro_2021 |> mutate(month = lubridate::month(month)))