---
title: "Fundamentals"
subtitle: "IQA Lecture 1"
author: "Charles Lanfear"
date: "12 Oct 2022<br>Updated: `r gsub(' 0', ' ', format(Sys.Date(), format='%d %b %Y'))`"
output:
  xaringan::moon_reader:
    css: [xaringan-themer.css, "../assets/cam_style.css"]
    lib_dir: libs
    nature:
      highlightStyle: tomorrow-night-bright
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "../assets/cam_macros.js"
      titleSlideClass: ["center","top"]
---

```{r setup}
#| include: false
knitr::opts_chunk$set(eval=TRUE, echo=TRUE, message=TRUE, warning=TRUE)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  white_color = "#f0f1eb",
  black_color = "#28282B",
  primary_color = "#003C71",
  secondary_color = "#8A1538",
  header_font_google = google_font("Montserrat"),
  text_font_google   = google_font("Open Sans"),
  code_font_google   = google_font("Fira Mono"),
  padding = "16px 32px 16px 32px",
  text_font_size = "1.3rem",
  code_font_size = "1.1rem"
)
```

# Today

* Math Review
   * Variables, vectors, and matrices
   * Lines and curves
   * Derivatives
* Data manipulation with `dplyr`

---
class: inverse
# Math Review

---
## Variables

**Variables** are symbols representing sets of one or more **elements** which might take any number of values.

* Letters like $x$, $y$, and $z$ are commonly used to indicate variables.

  * e.g., $x = 3$

--

* Capital letters ($X$) or letters with an index ($x_i$) refer to variables with multiple values---that is, with *dimensions*.
  * e.g.: $X = x_i = (3,4,5)$


--

* Subscripts like $x_i$ are used to **index** elements of vectors.

   * $i$ here is itself a variable that indicates the position

--

   * $x_1 = 3$, $x_2 = 4$, $x_3 = 5$



---
## Vectors in R

```{r}
x <- c(3, 4, 5) # Create x
x
```

--

Getting $x_3$:

```{r}
x[3] # Get the third element of x
```

--

We can index multiple elements.

Get $x_2$ and $x_3$:

```{r}
x[c(2,3)] # Get the second and third elements of x
```

---
## Matrices

**Matrices** are rectangular tables of numbers. They're typically indicated by a capital letter, e.g., $X$

&nbsp;

$X_{i,j} = \begin{bmatrix}x_{1,1} & x_{1,2}\\x_{2,1} & x_{2,2}\end{bmatrix}$

&nbsp;

Matrices are indexed with subscripts for *rows*, *columns* (e.g., $x_{ij}$)

--

&nbsp;

.pull-left[
$X = \begin{bmatrix}3 & 5\\4 & 6\end{bmatrix}$
]
.pull-right[
$x_{2,1} = 4$
]



---
## Matrices in R

```{r}
(X <- matrix(c(3,4,5,6), nrow = 2))
```

Note R shows indices on the margins to tell you how to subset.

```{r}
X[2,1]
```

---
## Summation

$\sum_{i=1}^{n}x_i$

"Sum all values of $x$ from the first ($i=1$) until the last ($n$)"

--

Given $x = [7, 11, 11, 13, 26]$

$x_1 + x_2 + x_3 + x_4 = 7 + 11 + + 11 + 13 + 26$

--

```{r}
x <- c(7, 11, 11, 13, 26)
sum(x)
```
Often when summing all elements of a vector, the sub/super scripts are hidden

* e.g. $\sum x_i = \sum_{i=1}^{n}x_i$

---
class: inverse

# Measures of Central Tendency

### Otherwise known as averages

---

## Mean

The mean is the *expected value* of a set of variable.

$\bar{x} = \frac{1}{n}\sum_{i=1}^{n}x_{i}$

--

That is, if you draw randomly from that variable, the mean would be the best guess.<sup>1</sup>

.footnote[[1] Technically the mean minimizes the *squared error* of guesses.]

--

```{r}
(1/length(x)) * sum(x)
```

```{r}
mean(x)
```


---

## Median

The **median** has this heinous formula:

$m(x_i) = \begin{cases} x_{\frac{n+1}{2}},& \text{if } n \text{ odd}\\ \frac{1}{2}(x_{\frac{n}{2}} + x_{\frac{n}{2} + 1}), &\text{if } n \text{ even}\end{cases}$


"If $x$ has an odd number of elements, the median is the middle one. If it has an even number of elements, average the middle two."

--

```{r}
x[(length(x) + 1) / 2]
```


```{r}
median(x)
```

---

## Mode

The **mode** is the most common value in the variable.

--

There are formulas for the mode, but they aren't very intuitive, despite it being the most intuitive measure of central tendency.

--

You can use a `table()` to see frequencies of values:

```{r}
table(x)
```

--

And you can find the mode directly:

```{r}
table(x)[table(x)==max(table(x))]
```

---
# Extreme Values

The mean is sensitive to extreme values:

```{r}
z <- c(2, 5, 3, 5, 95)
mean(z)
```

--

The median (and mode) are not:

```{r}
median(z)
```

--

This means the median may be a more useful "average" when your data have a small number of extreme values.

This is common with things like income or self-reported number of crimes committed.

---

# Measures of Dispersion

---

## Variance<sup>1</sup>

The variance measures how dispersed data are around the mean. Typically we use the *sample* variance:

$$
s^2 = \frac{\sum (x_i - \bar{x})^2}{n - 1}
$$

We'll see this in action next week when we look at distributions.

--

```{r}
(s2 <- sum((x - mean(x))^2) / (length(x) -1))
var(x)
```

--

If every value is the same, the variance is *zero*.

---

## Standard Deviation

The standard deviation ($s$ or $sd$) is just the root of the variance:

$s = sd = \sqrt{s^2}$

--

You can interpret it as the "typical" distance from values in the data from their mean.

--

```{r}
sqrt(var(x))
sd(x)
```

---
class: inverse
# Lines

---

# Slope-Intercept

Mathematically, lines can be defined by a slope and an intercept.

--

You've seen this before, perhaps many moons ago:

$y = mx + b$

--

We'll restate it this way:

$y = a + bx$

--

$a$ is the **intercept**

* The value of $y$ when $x = 0$

--

$b$ is the **slope**

* The units of $y$ the line changes for every unit of $x$

---

# Intercept


$y = 1 + 0.5x$


```{r}
#| fig.height: 4
plot(c(0,5), c(0,5), type = "n", xlab = "x", ylab = "y")
abline(a = 1, b = 0.5)
```


The line *intercepts* the y-axis at 1.

---
# Intercept

$y = 3 + 0.5x$

```{r}
#| fig.height: 4
plot(c(0,5), c(0,5), type = "n", xlab = "x", ylab = "y")
abline(a = 3, b = 0.5)
```

The line *intercepts* the y-axis at 3.

---
# Intercept

$y = 2 + 0.5x$

```{r}
#| fig.height: 4
plot(c(0,5), c(0,5), type = "n", xlab = "x", ylab = "y")
abline(a = 2, b = 0.5)
```

The line *intercepts* the y-axis at 2.

---
# Slope

$y = 2 + 0.5x$

```{r}
#| fig.height: 4
plot(c(0,5), c(0,5), type = "n", xlab = "x", ylab = "y")
abline(a = 2, b = 0.5)
```

From 2, the line increases by 0.5 for every $x$.

---
# Slope

$y = 2 + 0x$

```{r}
#| fig.height: 4
plot(c(0,5), c(0,5), type = "n", xlab = "x", ylab = "y")
abline(a = 2, b = 0)
```

From 2, the line increases by 0 for every $x$.

---
# Slope

$y = 2 + 2x$

```{r}
#| fig.height: 4
plot(c(0,5), c(0,5), type = "n", xlab = "x", ylab = "y")
abline(a = 2, b = 2)
```

From 2, the line increases by 2 for every $x$.

---
class: inverse

# Derivatives

**Derivatives** ( $\frac{dy}{dx}$ ) are related to slopes.

The derivative is the rate of change (the slope) at a given point of a function---the amount $y$ increases ( $dy$ ) for a unit of $x$ ( $dx$ ).

--

For a straight line, the slope is always the same and is equal to the derivative.

--

For curves, the slope is different depending on where on the curve you're looking.

---
class: inverse

## Polynomials

---
# Quadratics

You can define a curve by augmenting the slope-intercept formula:

$y = 2 + 0.5x + 0.25x^2$

--

A *squared* or **quadratic** term ($x^2$) creates a *parabola*.

```{r}
#| fig.height = 4
curve(2 + 0.5*x + 0.25*x^2, from = -2, to = 2, ylab = "y")
```

--

.pull-right-30[
.footnote[
What is the slope of this curve?
]
]

---
# Taking the Derivative

While a curve has many different slopes, all those slopes can be defined by a single derivative<sup>1</sup>

.footnote[[1] At least for any curves we're going to talk about!]

--

Given $f(x) = a + x^n$, $\frac{d}{dx}f(x) = nx^{n-1}$

Basic rules:

* Delete any terms without $x$ (e.g., $a=2$)
* Premultiply by exponents and decrement it

--

So...

* $y = 2 + 0.5x + 0.25x^2$
* $\frac{dy}{dx} = 0.5 + 0.5x$
* When $x = 3$ the slope is $0.5 + 0.5*3 = 2$.

---
# Okay, that's enough maths

---
class: inverse
# Oh, thank god

---
# Time for more code

---
class: inverse

.text-align-center[
# BOO
]

---
class: inverse

# Part II

## Data manipulation with `dplyr`

![](img/dplyr.svg)